Model Checking Differentially Private Properties

### 2018-01-28 applications.tex
use standard differential privacy definition e^{\epsilon}

\indent
\emph{Privacy analysis}.
In the processing phase, the bit $f_a$ for an event type $a$ can only
have two different distributions $U$ and $B$. If the event type is
never observed, $\Pr_U[f_a = 0] = \Pr_U[f_a = 1] = \frac{1}{2}$; if
the event type has been observed, we have $\Pr_B[f_a = 0] =
\frac{3}{8}$ and $\Pr_B[f_a = 1] = \frac{5}{8}$. Note that
$\frac{1}{2} \cdot \frac{3}{8} \leq \frac{1}{2} \leq 2 \cdot
\frac{3}{8}$ and $\frac{1}{2} \cdot \frac{5}{8} \leq \frac{1}{2} \leq
2 \cdot \frac{5}{8}$. For every $x \in \{ 0, 1 \}$, we have
$\frac{1}{2} \cdot \Pr_B[f_a = x] \leq \Pr_U[f_a = x] \leq 2 \cdot
\Pr_B[f_a = x]$. On neighboring observations, the distribution of
$f_a$ can only be changed from $U$ to $B$, values of $f_a$ is $(2,
0)$-differentially private and similarly for $f_b$. Finally, the
$1/2$-geometric mechanism gives $(2, 0)$-differential privacy. By
composition theorem, the steaming data analysis mechanism in
Algorithm~\ref{algorithm:frequency-estimator} is $(4,
0)$-differentially private.

@@

\noindent
\emph{Privacy Specification}.
For classical differential privacy, we are interested in the output
distribution. Algorithm~\ref{algorithm:frequency-estimator} has three
different outputs. By privacy analysis, the state $I$ satisfies the
following $\dpCTLstar$ formula:
\[
\dpriv{4}{0} \F 0 \wedge \dpriv{4}{0} \F 1 \wedge \dpriv{4}{0} \F 2
\]
For pan privacy, we moreover wish to measure privacy loss in internal
states. For observations of length $k$, the probabilities of reaching
each internal states need to be sufficiently close. The state $I$
therefore must satisfy
\[
\dpriv{4}{0} \F 0 \wedge \dpriv{4}{0} \F 1 \wedge \dpriv{4}{0} \F 2
\wedge
\bigwedge_{k \in \bbfN}
\dpriv{4}{0} \X^k 00 \wedge 
\dpriv{4}{0} \X^k 01 \wedge 
\dpriv{4}{0} \X^k 10 \wedge 
\dpriv{4}{0} \X^k 11.
\]

>>>>

\indent
\emph{Privacy analysis}.
In the processing phase, the bit $f_a$ for an event type $a$ can only
have two different distributions $U$ and $B$. If the event type is
never observed, $\Pr_U[f_a = 0] = \Pr_U[f_a = 1] = \frac{1}{2}$; if
the event type has been observed, we have $\Pr_B[f_a = 0] =
\frac{3}{8}$ and $\Pr_B[f_a = 1] = \frac{5}{8}$. Note that
$\frac{1}{2} \cdot \frac{3}{8} \leq \frac{1}{2} \leq 2 \cdot
\frac{3}{8}$ and $\frac{1}{2} \cdot \frac{5}{8} \leq \frac{1}{2} \leq
2 \cdot \frac{5}{8}$. For every $x \in \{ 0, 1 \}$, we have
$\frac{1}{2} \cdot \Pr_B[f_a = x] \leq \Pr_U[f_a = x] \leq 2 \cdot
\Pr_B[f_a = x]$. On neighboring observations, the distribution of
$f_a$ can only be changed from $U$ to $B$, values of $f_a$ is $(2,
0)$-differentially private and similarly for $f_b$. Finally, the
$1/2$-geometric mechanism gives $(\ln 2, 0)$-differential privacy. 
Recall $2\ln 2 = \ln 4$,
the steaming data analysis mechanism in
Algorithm~\ref{algorithm:frequency-estimator} is $(\ln 4,
0)$-differentially private by
composition theorem.

@@

\noindent
\emph{Privacy Specification}.
For classical differential privacy, we are interested in the output
distribution. Algorithm~\ref{algorithm:frequency-estimator} has three
different outputs. By privacy analysis, the state $I$ satisfies the
following $\dpCTLstar$ formula:
\[
\dpriv{\ln 4}{0} \F 0 \wedge \dpriv{\ln 4}{0} \F 1 \wedge 
\dpriv{\ln 4}{0} \F 2
\]
For pan privacy, we moreover wish to measure privacy loss in internal
states. For observations of length $k$, the probabilities of reaching
each internal states need to be sufficiently close. The state $I$
therefore must satisfy
\[
\dpriv{\ln 4}{0} \F 0 \wedge \dpriv{2\ln 4}{0} \F 1 \wedge 
\dpriv{\ln 4}{0} \F 2
\wedge
\bigwedge_{k \in \bbfN}
\dpriv{\ln 4}{0} \X^k 00 \wedge 
\dpriv{\ln 4}{0} \X^k 01 \wedge 
\dpriv{\ln 4}{0} \X^k 10 \wedge 
\dpriv{\ln 4}{0} \X^k 11.
\]

### 2018-01-28 examples.tex
mdps
use standard differential privacy definition e^{\epsilon}

\begin{example}\label{exa:survey}
  Figure~\ref{figure:2-dp} shows the survey
mechanism and its corresponding Markov chain. In the figure, the states
$+$ and $-$ denote the diagnoses are positive or negative
respectively; the states $Y$ and $N$ denote answers to the survey
question. We have $AP=\{Y,N\}$, $L(Y)=\{Y\}$, $L(N)=\{N\}$, and $L(+)=L(-)=\emptyset$. States $+$ and $-$ are neighbors.\lz{draw it on the picture?}

@@

Let us recall briefly the analysis of the survey mechanism. The data universe
$\mathcal{X}$ is $\{ +, - \}$. The mechanism $M$ is a randomized
algorithm with inputs in $\mathcal{X}$ and outputs in $\{ Y, N
\}$. For any $x \in \mathcal{X}$, we have $1/4 \leq \Pr[M(x) = Y] \leq
3/4$. Hence for any neighboring $x, x' \in \mathcal{X}$, $\Pr[M(x) =
Y] \leq 3/4 = 3 \cdot 1/4 \leq 3 \Pr[M (x') =
Y]$. Similarly, $\Pr[M (x) = N] \leq 3 \Pr[M (x') = N]$ for any
neighboring $x, x' \in \mathcal{X}$. The survey mechanism is
$(3, 0)$-differentially private.

\paragraph{Truncated $\alpha$-Geometric Mechanism.}
More sophisticated differentially private mechanisms are
available. Consider a query
$f : \mathcal{X}^n \rightarrow \bbfZ^{\geq 0}$. Let $\alpha \in (0, 1)$.
The \emph{$\alpha$-geometric mechanism}
outputs $f(x) + Y$ where $Y$ is a random variable with the geometric
distribution~\cite{GRS:09:UUPM,GRS:12:UUPM} :
\[
\Pr[Y = y] = \frac{1 - \alpha}{1 + \alpha}\alpha^{|y|}
\textmd{ for } y \in \bbfZ.
\]
The $\alpha$-geometric mechanism is oblivious since it has the same
output distribution on any inputs $x, x'$ with $f (x) = f
(x')$. Moreover, it is also $(\alpha^{\Delta (f)}, 0)$-differentially
private for any query $f$ with sensitivity $\Delta (f)$. The range of
the mechanism
however is $\bbfZ$. It may give nonsensical outputs such as
negative integers for queries with positive ranges.

The \emph{truncated $\alpha$-geometric mechanism over $\underline{n}$}
outputs $f (x) + Z$ where $Z$ is a random variable with the following
distribution:
\[
\Pr[Z = z] =
\left\{
  \begin{array}{ll}
    0 & \textmd{ if } z < - f (x) \\
    \frac{\alpha^{f (x)}}{1 + \alpha} & \textmd{ if } z = -f (x)\\
    \frac{1 - \alpha}{1 + \alpha}\alpha^{|z|} &
    \textmd{ if } -f (x) < z < n - f (x)\\
    \frac{\alpha^{n - f (x)}}{1 + \alpha} & \textmd{ if } z = n-f (x)\\
    0 & \textmd{ if } z > n - f (x)
  \end{array}
\right.
\]
Note the range of the truncated $\alpha$-geometric mechanism is
$\underline{n}$. The truncated $\alpha$-geometric mechanism is
again oblivious; it is also $(\alpha^{\Delta (f)}, 0)$-differentially
private for any query $f$ with sensitivity $\Delta (f)$.
The truncated $1/2$-geometric mechanism over $\underline{5}$ is
given in Figure~\ref{figure:geometric-mechanism-table}.

>>>>

\begin{example}\label{exa:survey}
  Figure~\ref{figure:2-dp} shows the survey
mechanism and its corresponding Markov chain. In the figure, the states
$+$ and $-$ denote the diagnoses are positive or negative
respectively; the states $Y$ and $N$ denote answers to the survey
question. We have $AP=\{Y,N\}$, $L(Y)=\{Y\}$, $L(N)=\{N\}$, and $L(+)=L(-)=\emptyset$. States $+$ and $-$ are neighbors.\lz{In general, we omit a special state $\dagger$ with $L(\dagger)=\emptyset$, and all omitted transitions in the picture lead to this state.}

@@

Let us recall briefly the analysis of the survey mechanism. The data universe
$\mathcal{X}$ is $\{ +, - \}$. The mechanism $M$ is a randomized
algorithm with inputs in $\mathcal{X}$ and outputs in $\{ Y, N
\}$. For any $x \in \mathcal{X}$, we have $1/4 \leq \Pr[M(x) = Y] \leq
3/4$. Hence for any neighboring $x, x' \in \mathcal{X}$, $\Pr[M(x) =
Y] \leq 3/4 = 3 \cdot 1/4 \leq 3 \Pr[M (x') =
Y]$. Similarly, $\Pr[M (x) = N] \leq 3 \Pr[M (x') = N]$ for any
neighboring $x, x' \in \mathcal{X}$. The survey mechanism is
$(\ln 3, 0)$-differentially private.

\paragraph{Truncated $\alpha$-Geometric Mechanism.}
More sophisticated differentially private mechanisms are
available. Consider a query
$f : \mathcal{X}^n \rightarrow \bbfZ^{\geq 0}$. Let $\alpha \in (0, 1)$.
The \emph{$\alpha$-geometric mechanism}
outputs $f(x) + Y$ where $Y$ is a random variable with the geometric
distribution~\cite{GRS:09:UUPM,GRS:12:UUPM} :
\[
\Pr[Y = y] = \frac{1 - \alpha}{1 + \alpha}\alpha^{|y|}
\textmd{ for } y \in \bbfZ.
\]
The $\alpha$-geometric mechanism is oblivious since it has the same
output distribution on any inputs $x, x'$ with $f (x) = f
(x')$. Moreover, it is also $(- {\Delta (f)} \ln \alpha, 0)$-differentially
private for any query $f$ with sensitivity $\Delta (f)$. The range of
the mechanism
however is $\bbfZ$. It may give nonsensical outputs such as
negative integers for queries with positive ranges.

The \emph{truncated $\alpha$-geometric mechanism over $\underline{n}$}
outputs $f (x) + Z$ where $Z$ is a random variable with the following
distribution:
\[
\Pr[Z = z] =
\left\{
  \begin{array}{ll}
    0 & \textmd{ if } z < - f (x) \\
    \frac{\alpha^{f (x)}}{1 + \alpha} & \textmd{ if } z = -f (x)\\
    \frac{1 - \alpha}{1 + \alpha}\alpha^{|z|} &
    \textmd{ if } -f (x) < z < n - f (x)\\
    \frac{\alpha^{n - f (x)}}{1 + \alpha} & \textmd{ if } z = n-f (x)\\
    0 & \textmd{ if } z > n - f (x)
  \end{array}
\right.
\]
Note the range of the truncated $\alpha$-geometric mechanism is
$\underline{n}$. The truncated $\alpha$-geometric mechanism is
again oblivious; it is also $(- {\Delta (f)} \ln \alpha, 0)$-differentially
private for any query $f$ with sensitivity $\Delta (f)$.
The truncated $1/2$-geometric mechanism over $\underline{5}$ is
given in Figure~\ref{figure:geometric-mechanism-table}.

### 2018-01-28 mdp.tex
use standard differential privacy definition e^{\epsilon}

In the figure, the states $+$,
$-$, $Y$, and $N$ are interpreted as before. The actions $L$ and $H$
denote low- and high-accuracy requests respectively. Let $M_H$ denote
the Markov chain derived by high-accuracy requests.
Observe that $1/5 \leq \Pr[M_H(x) = Y] \leq 4/5$ for any $x
\in \mathcal{X}$. Hence $\Pr[M_H (x) = Y] \leq 4/5 = 4 \cdot 1/5 \leq
4 \Pr[M_H (x') = Y]$ for any neighboring $x, x' \in
\mathcal{X}$. Similarly, $\Pr[M_H (x) = N] \leq 4 \Pr[M_H (x') =
N]$. The high-accuracy mechanism is $(4,0)$-differentially private.
The privacy guarantees vary from accuracy requests.

\subsection{Semantics}
The semantics over Markov chains generalizes to Markov decision
processes as well. Note that two states with different enabled actions
are trivially distinguishable; no privacy can be preserved. Without
loss of generality, we consider reactive Markov decision processes.
Let $M = (S, \Act, \wp, L)$ be a reactive Markov
decision process. Define the satisfaction
relation $M, \neighbor{S}, s \models \Phi$ as follows.
\begin{eqnarray*}
  M, \neighbor{S}, s \models \top\\
  M, \neighbor{S}, s \models p
  & \textmd{ if } &
  p \in L(s)\\
  M, \neighbor{S}, s \models \neg \Phi
  & \textmd{ if } &
  M, \neighbor{S}, s \not\models \Phi\\
  M, \neighbor{S}, s \models \Phi_0 \wedge \Phi_1
  & \textmd{ if } &
  M, \neighbor{S}, s \models \Phi_0 \textmd{ and }
  M, \neighbor{S}, s \models \Phi_1\\
  M, \neighbor{S}, s \models \PJ{J} \phi
  & \textmd{ if } &
  \myPr{s}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} \in J
  \textmd{ for every scheduler } \scheduler{S}\\
  M, \neighbor{S}, s \models \dpriv{\epsilon}{\delta} \phi
  & \textmd{ if } &
  \textmd{for all } t \textmd{ with } s \neighbor{S} t \textmd{ and
   scheduler } \\
  & & \scheduler{S},
   \myPr{s}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} \leq \epsilon \cdot
   \myPr{t}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} + \delta
   \textmd{ and }
   \myPr{t}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} \leq \epsilon \cdot
  \\
  & &
   \myPr{s}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} + \delta
\end{eqnarray*}

@@

\noindent
\emph{Discussion.}
\lz{this part not needed?}
Another definition of the differentially private operator
might be:
\begin{eqnarray*}
  M, \neighbor{S}, s \models \dpriv{\epsilon}{\delta}^{\mathit{bad}} \phi
  & \textmd{ if } &
  \textmd{for all } t \textmd{ with } s \neighbor{S} t \textmd{ and
  scheduler } \scheduler{S},
  \myPr{s}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} \leq
  \epsilon \cdot\\
  && \myPr{t}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} + \delta \textmd{ and }
  \myPr{t}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} \leq
  \epsilon \cdot \myPr{s}{M_{\scheduler{S}}}{\neighbor{S}}{\phi}
  + \delta
\end{eqnarray*}
A state satisfies $\dpriv{\epsilon}{\delta}^{\mathit{bad}} \phi$ if
no scheduler could differentiate the probabilities of paths satisfying
$\phi$ from neighbors. Such a definition would allow strategies to
take different actions from different states. It might need different
action sequences to distinguish neighbors. Two neighbors could still be
differentiated by a action sequence and lose their privacy. We hence
prefer the original definition.

@@

Our next task is to compute $\{ M, \neighbor{S}, s \models
\dpriv{\epsilon}{\delta} \phi \}$ for a Markov decision process $M =
(S, \Act, \wp, L)$. Recall the semantics of $\dpriv{\epsilon}{\delta}
\phi$. Given $s, t$ with $s \neighbor{M} t$ and a path formula
$\phi$, we need to decide whether
$\myPr{s}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} \leq
\epsilon \myPr{t}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} + \delta$
for every scheduler $\scheduler{S}$.
If the path formula only contain next operators, it is easy to deal with. For $\phi:=\X B$ with $B\subseteq S$ only one step behaviour needs to be considered, thus only the first action in the query sequence is needed to compute the maximum
\begin{eqnarray}
  \label{eqn:max-nextB}
\max_{\scheduler{S}}
\myPr{s}{M_{\scheduler{S}}}{\neighbor{S}}{\gX B} -
\epsilon \myPr{t}{M_{\scheduler{S}}}{\neighbor{S}}{\gX B}
\end{eqnarray}
with $s, t \in S$. It can be easily generalized to nested next operators:  one needs only to enumerate
all actions query sequences of length obtained by the number of nexted $\X$ operators.

>>>>

In the figure, the states $+$,
$-$, $Y$, and $N$ are interpreted as before. The actions $L$ and $H$
denote low- and high-accuracy requests respectively. Let $M_H$ denote
the Markov chain derived by high-accuracy requests.
Observe that $1/5 \leq \Pr[M_H(x) = Y] \leq 4/5$ for any $x
\in \mathcal{X}$. Hence $\Pr[M_H (x) = Y] \leq 4/5 = 4 \cdot 1/5 \leq
4 \Pr[M_H (x') = Y]$ for any neighboring $x, x' \in
\mathcal{X}$. Similarly, $\Pr[M_H (x) = N] \leq 4 \Pr[M_H (x') =
N]$. The high-accuracy mechanism is $(\ln 4,0)$-differentially private.
The privacy guarantees vary from accuracy requests.

\subsection{Semantics}
The semantics over Markov chains generalizes to Markov decision
processes as well. Note that two states with different enabled actions
are trivially distinguishable; no privacy can be preserved. Without
loss of generality, we consider reactive Markov decision processes.
Let $M = (S, \Act, \wp, L)$ be a reactive Markov
decision process. Define the satisfaction
relation $M, \neighbor{S}, s \models \Phi$ as follows.
\begin{eqnarray*}
  M, \neighbor{S}, s \models \top\\
  M, \neighbor{S}, s \models p
  & \textmd{ if } &
  p \in L(s)\\
  M, \neighbor{S}, s \models \neg \Phi
  & \textmd{ if } &
  M, \neighbor{S}, s \not\models \Phi\\
  M, \neighbor{S}, s \models \Phi_0 \wedge \Phi_1
  & \textmd{ if } &
  M, \neighbor{S}, s \models \Phi_0 \textmd{ and }
  M, \neighbor{S}, s \models \Phi_1\\
  M, \neighbor{S}, s \models \PJ{J} \phi
  & \textmd{ if } &
  \myPr{s}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} \in J
  \textmd{ for every scheduler } \scheduler{S}\\
  M, \neighbor{S}, s \models \dpriv{\epsilon}{\delta} \phi
  & \textmd{ if } &
  \textmd{for all } t \textmd{ with } s \neighbor{S} t \textmd{ and
   scheduler } \scheduler{S}, 
   \myPr{s}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} \leq 
  e^{\epsilon} \cdot
   \\
  & & 
   \myPr{t}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} + \delta
   \textmd{ and }
   \myPr{t}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} \leq e^{\epsilon} \cdot
   \myPr{s}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} + \delta
\end{eqnarray*}

@@

\noindent
\emph{Discussion.}
\lz{this part not needed?}
Another definition of the differentially private operator
might be:
\begin{eqnarray*}
  M, \neighbor{S}, s \models \dpriv{\epsilon}{\delta}^{\mathit{bad}} \phi
  & \textmd{ if } &
  \textmd{for all } t \textmd{ with } s \neighbor{S} t \textmd{ and
  scheduler } \scheduler{S},
  \myPr{s}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} \leq
  e^{\epsilon} \cdot\\
  && \myPr{t}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} + \delta \textmd{ and }
  \myPr{t}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} \leq
  e^{\epsilon} \cdot \myPr{s}{M_{\scheduler{S}}}{\neighbor{S}}{\phi}
  + \delta
\end{eqnarray*}
A state satisfies $\dpriv{\epsilon}{\delta}^{\mathit{bad}} \phi$ if
no scheduler could differentiate the probabilities of paths satisfying
$\phi$ from neighbors. Such a definition would allow strategies to
take different actions from different states. It might need different
action sequences to distinguish neighbors. Two neighbors could still be
differentiated by a action sequence and lose their privacy. We hence
prefer the original definition.

@@

Our next task is to compute $\{ M, \neighbor{S}, s \models
\dpriv{\epsilon}{\delta} \phi \}$ for a Markov decision process $M =
(S, \Act, \wp, L)$. Recall the semantics of $\dpriv{\epsilon}{\delta}
\phi$. Given $s, t$ with $s \neighbor{M} t$ and a path formula
$\phi$, we need to decide whether
$\myPr{s}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} \leq
e^{\epsilon} \myPr{t}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} + \delta$
for every scheduler $\scheduler{S}$.
If the path formula only contain next operators, it is easy to deal with. For $\phi:=\X B$ with $B\subseteq S$ only one step behaviour needs to be considered, thus only the first action in the query sequence is needed to compute the maximum
\begin{eqnarray}
  \label{eqn:max-nextB}
\max_{\scheduler{S}}
\myPr{s}{M_{\scheduler{S}}}{\neighbor{S}}{\gX B} -
e^{\epsilon} \myPr{t}{M_{\scheduler{S}}}{\neighbor{S}}{\gX B}
\end{eqnarray}
with $s, t \in S$. It can be easily generalized to nested next operators:  one needs only to enumerate
all actions query sequences of length obtained by the number of nexted $\X$ operators.

### 2018-01-28 specifications.tex
use standard differential privacy definition e^{\epsilon}

 Define the $\dpCTL$ formula
$\psi = \dpriv{.5}{0} (\F out_0) \wedge \dpriv{.5}{0} (\F out_1) \wedge
\cdots \wedge \dpriv{.5}{0} (\F out_5)$\lz{I think we better use $\X$ operator here ;)}. If the state $in_k$ satisfies
$\psi$ for $k = 0, \ldots, 5$, then the $.5$-geometric mechanism is
$(.5, 0)$-differentially private.

\paragraph{Compositional Properties.}
Compositional aspect is one of the building block for differentially private algorithms. According to the compositional theorem \cite[Theorem 3.16]{DR:14:AFDP}, for $(\epsilon_1,\delta_1)$-differentially private algorithm $M_1$ and
$(\epsilon_2,\delta_2)$-differentially private algorithm $M_2$, their combination $M$ defined by $M(x)=(M_1(x), M_2(x))$ is $(\epsilon_1\epsilon_2,\delta_1+\delta_2)$-differentially private. The degradation is rooted in the repeated computation. To illustrate this property, we consider the extended survey example below where the curator allows two consequent queries.
The corresponding model is depicted in Figure~\ref{figure:2-dp-mc}, where states $Y_1,Y_2$ are labelled by $Y$, and $N_1,N_2$ are labelled by $N$.

@@

We consider the formula $\dpriv{9}{0} (\X (Y\wedge \X Y))$. A path satisfies $\X (Y\wedge \X Y)$ if the second state satisfies $Y$ and the third state satisfies $Y$ as well. We verify that this formula is satisfied  for state $+$ and $-$. Moreover, the bound $\epsilon=9$ is tight, since the probability of satisfying $\X (Y\wedge \X Y)$ from states $+$ and $-$ are $\frac{9}{16}$ and $\frac{1}{16}$, respectively. Finally, the formula $\wedge_{a_1,a_2}\dpriv{9}{0} (\X (a_1\wedge \X a_2))$ corresponds to the desired compositional property for the model, where $a_1,a_2$ range over all atomic propositions $\{Y,N\}$.

We consider two other formulae for comparision: 
(a) Formula $\dpriv{3}{0} (\X \X Y)$. In this case we donot have privacy lost, even though we are querying twice: the reason is that the output of the first query is not utilized at all. It is easy to verify that it is indeed satisfied by $+$ and $-$.
(b) Formula $\dpriv{3}{0} (\X (Y\wedge \dpriv{3}{0}(\X Y)))$. This is a nested $\dpCTL$ formula, where the inner state formula $\dpriv{3}{0}(\X Y)$ states the one step differential privacy. Observe the inner formula is satisfied by all neighbours. Thus, the outer formula also has no privacy lost.

\paragraph{Tighter Privacy Bounds for Compositional Properties.}
One of the advantage of applying model checking is that we may get tighter bounds for compositional properties. Consider our survey example, and the property $\dpriv{1}{.5} (\X Y)$. Obviously, it holds in states $+$ and $-$. A careful check ensures that one cannot decrease the value of $\epsilon_1=1$ or $\delta_1=.5$ without increasing the other one.

Now we consider the formula $\dpriv{\epsilon_2}{\delta_2} (\X (Y\wedge \X Y))$ under the model in Figure~\ref{figure:2-dp-mc}. Applying the compositional theorem, one can choose $\epsilon_2=\epsilon_1^2=1$ and $\delta_2=2\delta_1=1$. However, we can check easily that one gets better privacy parameter $(1,0.5)$ or $(0,\frac{9}{16})$.

>>>>

 Define the $\dpCTL$ formula
$\psi = \dpriv{.5}{0} (\X out_0) \wedge \dpriv{.5}{0} (\X out_1) \wedge
\cdots \wedge \dpriv{.5}{0} (\X out_5)$. If the state $in_k$ satisfies
$\psi$ for $k = 0, \ldots, 5$, then the $.5$-geometric mechanism is
$(\ln 2, 0)$-differentially private.

\paragraph{Compositional Properties.}
Compositional aspect is one of the building block for differentially private algorithms. According to the compositional theorem \cite[Theorem 3.16]{DR:14:AFDP}, for $(\epsilon_1,\delta_1)$-differentially private algorithm $M_1$ and
$(\epsilon_2,\delta_2)$-differentially private algorithm $M_2$, their combination $M$ defined by $M(x)=(M_1(x), M_2(x))$ is $(\epsilon_1 + \epsilon_2,\delta_1+\delta_2)$-differentially private. The degradation is rooted in the repeated computation. To illustrate this property, we consider the extended survey example below where the curator allows two consequent queries.
The corresponding model is depicted in Figure~\ref{figure:2-dp-mc}, where states $Y_1,Y_2$ are labelled by $Y$, and $N_1,N_2$ are labelled by $N$.

@@

We consider the formula $\dpriv{\ln 9}{0} (\X (Y\wedge \X Y))$. A path satisfies $\X (Y\wedge \X Y)$ if the second state satisfies $Y$ and the third state satisfies $Y$ as well. We verify that this formula is satisfied  for state $+$ and $-$. Moreover, the bound $\epsilon=\ln 9$ is tight, since the probability of satisfying $\X (Y\wedge \X Y)$ from states $+$ and $-$ are $\frac{9}{16}$ and $\frac{1}{16}$, respectively. Finally, the formula $\wedge_{a_1,a_2}\dpriv{\ln 9}{0} (\X (a_1\wedge \X a_2))$ corresponds to the desired compositional property for the model, where $a_1,a_2$ range over all atomic propositions $\{Y,N\}$.

We consider two other formulae for comparision: 
(a) Formula $\dpriv{\ln 3}{0} (\X \X Y)$. In this case we donot have privacy lost, even though we are querying twice: the reason is that the output of the first query is not utilized at all. It is easy to verify that it is indeed satisfied by $+$ and $-$.
(b) Formula $\dpriv{\ln 3}{0} (\X (Y\wedge \dpriv{\ln 3}{0}(\X Y)))$. This is a nested $\dpCTL$ formula, where the inner state formula $\dpriv{\ln 3}{0}(\X Y)$ states the one step differential privacy. Observe the inner formula is satisfied by all neighbours. Thus, the outer formula also has no privacy lost.

\paragraph{Tighter Privacy Bounds for Compositional Properties.}
One of the advantage of applying model checking is that we may get tighter bounds for compositional properties. Consider our survey example, and the property $\dpriv{0}{.5} (\X Y)$. Obviously, it holds in states $+$ and $-$. A careful check ensures that one cannot decrease the value of $\epsilon_1=0$ or $\delta_1=.5$ without increasing the other one.

Now we consider the formula $\dpriv{\epsilon_2}{\delta_2} (\X (Y\wedge \X Y))$ under the model in Figure~\ref{figure:2-dp-mc}. Applying the compositional theorem, one can choose $\epsilon_2=2\epsilon_1=0$ and $\delta_2=2\delta_1=1$. However, we can check easily that one gets better privacy parameter $(0,0.5)$ or $(-\infty,\frac{9}{16})$. \bw{use $-\infty$ instead of $\ln 0$. is it okay?}
